Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-07-23 Master Thesis/synopsis.tex
Consistent citations

\begin{proof}
See \cite{EP05} 3.4.1 and 3.4.3.
\end{proof}

@@

\begin{proof}
See \cite{EP05} 2.3, 4.1.4, \cite{ZS60} VI §3 Theorem 3.
\end{proof}

@@

\begin{proof}
By Lemma \ref{2.2-pre}, $w$ induces a valuation $\tilde{w}\in\operatorname{Spv}k(v)$. Let $k(v)_{\tilde{w}}$ be henselization of $k(v)$ with respect to $\tilde{w}$. By Theorem \ref{theorem-absoluteinertia}, $k(v)_{\tilde{w}}$ corresponds to an intermediate extension $K_v\subset L \subset K^\text{t}_v$. Let $v_L$ be the unique extension of $v$ on $L$. Then $k(v_L)=k(v)_{\tilde{w}}$. Let $w_L$ be the valuation on $L$ corresponding to $\tilde{w}\in \operatorname{Spv} k(v_L)$. By \cite{EP05} 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
\end{proof}

@@

\begin{proof}
(i) and (ii) follow from \cite{HK94}, Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} 18.6.14.
\end{proof}

@@

\begin{proof}
We adopt the notation from the previous proof. First, we shall prove that the following map is injective for every model $\mathfrak{X}$ of $K$:
\[ f: \mathrm{H}^3(K) \longrightarrow \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
For $x\in\mathfrak{X}$ let $\mathcal{O}_x$ be a henselization of $\mathcal{O}_{\mathfrak{X}, x}$ and $K_x$ its quotient field. For $y\in\mathfrak{X}_0$ set $P_y=\operatorname{Spec}(\mathcal{O}_y)_1$. For $z\in P_y$ let $\mathcal{O}_z$ be a henselization of $\mathcal{O}_y$ at $z$ and $K_z$ its quotient field. For $y\in\mathfrak{X}_0$ and $x\in\mathfrak{X}_1$ let $P_{x,y}$ be the subset of $P_y$ consisting of all $z\in P_y$ which are mapped to $x$ under $\operatorname{Spec}(\mathcal{O}_y)\to\mathfrak{X}$. Consider the following commutative diagram:
\[ \begin{tikzcd}
\mathrm{H}^3(K)\ar[r, "\varphi"]\ar[d, "f"'] & \displaystyle\prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_x)\ar[d, "\psi"]\\
\displaystyle\prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \ar[r, "g"']& \displaystyle\prod_{\substack{x\in\mathfrak{X}_1 \\ y\in\mathfrak{X}_0}}\prod_{z\in P_{x,y}}\mathrm{H}^3(K_z)
\end{tikzcd} \]
According to \cite{Sa87}, Theorem 5.3, $g$ is injective. Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma. Thus its image is contained in $\bigoplus\mathrm{H}^3(K_w)$. Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$. Now, Lemma \ref{2.3} (i) yields:
\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
Note that the right set is closed in $\operatorname{Spv}(K)$.
\end{proof}

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$. However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64}, Ch. I, §4, Proposition 22 we have
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64}, Ch. I, §4, Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72}, Theorem 20.21):
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$. By using this equation, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) is clear. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 0 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}

@@

\begin{proof}
See \cite{ZS60} VI §10 Theorem 18.
\end{proof}

>>>>

\begin{proof}
See \cite{EP05} Proposition 3.4.1 and Theorem 3.4.3.
\end{proof}

@@

\begin{proof}
See \cite{EP05} Section 2.3, Corollary 4.1.4, \cite{ZS60} VI, §3 Theorem 3.
\end{proof}

@@

\begin{proof}
By Lemma \ref{2.2-pre}, $w$ induces a valuation $\tilde{w}\in\operatorname{Spv}k(v)$. Let $k(v)_{\tilde{w}}$ be henselization of $k(v)$ with respect to $\tilde{w}$. By Theorem \ref{theorem-absoluteinertia}, $k(v)_{\tilde{w}}$ corresponds to an intermediate extension $K_v\subset L \subset K^\text{t}_v$. Let $v_L$ be the unique extension of $v$ on $L$. Then $k(v_L)=k(v)_{\tilde{w}}$. Let $w_L$ be the valuation on $L$ corresponding to $\tilde{w}\in \operatorname{Spv} k(v_L)$. By \cite{EP05} Corollary 4.1.4, $w_L$ is henselian iff both $v_L$ and $\tilde{w}$ are henselian, thus $L$ is a henselization of $w$.
\end{proof}

@@

\begin{proof}
(i) and (ii) follow from \cite{HK94} Lemma 2.4.3, and (iii) is a consequence of (ii) and the fact that henselizations commute with filtrated direct limits, e.g. see \cite{EGA4} Proposition 18.6.14.
\end{proof}

@@

\begin{proof}
We adopt the notation from the previous proof. First, we shall prove that the following map is injective for every model $\mathfrak{X}$ of $K$:
\[ f: \mathrm{H}^3(K) \longrightarrow \prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \]
For $x\in\mathfrak{X}$ let $\mathcal{O}_x$ be a henselization of $\mathcal{O}_{\mathfrak{X}, x}$ and $K_x$ its quotient field. For $y\in\mathfrak{X}_0$ set $P_y=\operatorname{Spec}(\mathcal{O}_y)_1$. For $z\in P_y$ let $\mathcal{O}_z$ be a henselization of $\mathcal{O}_y$ at $z$ and $K_z$ its quotient field. For $y\in\mathfrak{X}_0$ and $x\in\mathfrak{X}_1$ let $P_{x,y}$ be the subset of $P_y$ consisting of all $z\in P_y$ which are mapped to $x$ under $\operatorname{Spec}(\mathcal{O}_y)\to\mathfrak{X}$. Consider the following commutative diagram:
\[ \begin{tikzcd}
\mathrm{H}^3(K)\ar[r, "\varphi"]\ar[d, "f"'] & \displaystyle\prod_{x\in\mathfrak{X}_1}\mathrm{H}^3(K_x)\ar[d, "\psi"]\\
\displaystyle\prod_{y\in\mathfrak{X}_0}\mathrm{H}^3(K_y) \ar[r, "g"']& \displaystyle\prod_{\substack{x\in\mathfrak{X}_1 \\ y\in\mathfrak{X}_0}}\prod_{z\in P_{x,y}}\mathrm{H}^3(K_z)
\end{tikzcd} \]
According to \cite{Sa87} Theorem 5.3, $g$ is injective. Now, $\psi\circ\varphi$ can be identified with the injective map $\Phi$ from the proof of the previous Lemma. Thus its image is contained in $\bigoplus\mathrm{H}^3(K_w)$. Therefore, $f$ is injective as well and its image is contained in $\bigoplus\mathrm{H}^3(K_y)$. Now, Lemma \ref{2.3} (i) yields:
\[ \operatorname{supp}(f) = \bigcap_{\mathfrak{X}\in \mathfrak{M}(K)} \pi_\mathfrak{X}^{-1}(\{ y\in\mathfrak{X}_0\mid \operatorname{res}_{K_y/K}(f)\neq 0 \}) \]
Note that the right set is closed in $\operatorname{Spv}(K)$.
\end{proof}

@@

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = (G_K:G_\ell)$. However, $(G_K:G_\ell)$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64} Ch. I, §4 Proposition 22 we have
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ iff $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64} Ch. I, §4 Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

The second assertion follows from the first one, applied to $L$, by inspecting the following equation of supernatural numbers (\cite{En72} Theorem 20.21):
\[ [L:K] = (\Gamma':\Gamma) \cdot [k':k]\cdot p^d \]
for some $d \leq \infty$, where $\Gamma'$ denotes the value group and $k'$ the residue field of the unique prolongation of $w$ to $L$. By using this equation, it suffices to show the following equivalences:
\begin{enumerate}[(i)]
\item $\ell^\infty \nmid [k':k] \iff \ell^\infty \nmid [k':\mathbb{F}_p]$
\item $\ell^\infty\nmid (\Gamma':\Gamma) \iff \dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')=2$
\end{enumerate}
(i) is clear. For (ii) let $T'$ be the absolute inertia group of $L$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T'$ is isomorphic to $\mathbb{Z}_\ell^{r'}$, where $r'=\dim_{\mathbb{F}_\ell}(\Gamma'/\ell\Gamma')$. Furthermore, we have $(\Gamma':\Gamma) = (T:T')$ and the following exact sequence of $\ell$-Sylow groups:
\[ 0 \longrightarrow \mathbb{Z}_\ell^{r'} \longrightarrow \mathbb{Z}_\ell^{r}\longrightarrow (T/T')_\ell\longrightarrow 0 \]
Now $(T/T')_\ell$ is finite if and only if $r'=r=2$.
\end{proof}

@@

\begin{proof}
See \cite{ZS60} VI, §10 Theorem 18.
\end{proof}
